---
title: "hw03"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
**This is a high-level exploration on `gapminder` dataset. I would try to implemente the tasks on the Task Menu.**
-------------------
As usual, load the data first.
```{r}
suppressPackageStartupMessages(library(gapminder))
suppressPackageStartupMessages(library(tidyverse))
```

Then, go for the tasks below.

- **Get the maximum and minimum of GDP per capita for all continents.**

```{r}
df <- gapminder %>% 
  # for each continent
  group_by(continent) %>%  
  summarize(max = max(gdpPercap), min = min(gdpPercap))
```

```{r The max and min of GDP per capita for all continents}
dfplt <- df %>% 
  ggplot(aes(continent)) + 
  # plot the points in the same graph and differentiate by color
  geom_point(aes(y = max, colour = "max")) + 
  geom_point(aes(y = min, colour = "min")) +  
  ggtitle("The max and min of GDP per capita for all continents")
suppressPackageStartupMessages(library(gridExtra))
grid.arrange(tableGrob(df), dfplt, nrow = 1)
```

An interesting thing is that Asia has the highest and the lowest GDP per capita among all continents. 

- **Look at the spread of GDP per capita within the continents.**

```{r}
gapminder %>%
  group_by(continent) %>%
  summarize(mean = mean(gdpPercap), sd = sd(gdpPercap)) %>% 
  knitr::kable()
```


```{r The spread of GDP per capita within the continents}
gapminder %>% 
  ggplot(aes(x = gdpPercap, color = continent)) +
  scale_x_log10() +
  # show the distribution of data
  geom_density() + 
  ggtitle("The spread of GDP per capita within the continents")
```

The graph indicates the same conclusion with the first part: Asia has a hugh variance of GDP per capita and Oceania has a centralized distribution.  

- **Compute a trimmed mean of life expectancy for different years. Or a weighted mean, weighting by population. Just try something other than the plain vanilla mean.**

Compute the trimmed mean which discard the 10% largest and lowest value.

<div class="twoC">
```{r results = 'asis', echo=FALSE}
df <- gapminder %>% 
  group_by(year) %>% 
  summarize(trimmed_mean = mean(lifeExp,trim = 0.1))

knitr::kable(df)
```

```{r The trimmed mean of life expectancy over year, echo=FALSE}
df %>% 
  ggplot(aes(x = year,y = trimmed_mean))+
  geom_line() + 
  geom_point() +
  ggtitle("The trimmed mean of life expectancy over year")
```
</div>
<div class="twoC"></div>


In general, life expectancy is constantly increase these years.

- **How is life expectancy changing over time on different continents?**

Use the mean of life expectancy in every continent to represent overall situation.
```{r}
df <- gapminder %>% 
  group_by(continent,year) %>% 
  summarise(mean = mean(lifeExp))
knitr::kable(df)
```

```{r The mean of life expectancy over time on diffecent continents}
df %>% 
  ggplot(aes(y = mean,x = year)) +
  facet_wrap(~ continent) +
  geom_line() +
  geom_point() +
  ggtitle("The mean of life expectancy over time on diffecent continents")
```

- **Report the absolute and/or relative abundance of countries with low life expectancy over time by continent: Compute some measure of worldwide life expectancy – you decide – a mean or median or some other quantile or perhaps your current age. Then determine how many countries on each continent have a life expectancy less than this benchmark, for each year.**

Use the median of life expectancy as the worldwide benchmark.

```{r}
df <- gapminder %>% 
  # low_life means that lifeExp is less than the median
  mutate(low_life = lifeExp < median(lifeExp)) %>% 
  group_by(continent,year) %>% 
  # count the number of countries that satisfy the contidition
  summarise(num = sum(low_life==TRUE)) 

knitr::kable(df)
```

```{r Compare the life expectancy with the worldwide median}
gapminder %>% 
  ggplot(aes(x = continent, y = lifeExp)) +
  geom_jitter(aes(color = lifeExp < median(lifeExp), alpha = 0.8, shape = continent)) +
  facet_wrap(~ year) + 
  ggtitle("Compare the life expectancy with the worldwide median")

```

Scatter plot was used here to describe the life expectancy of each continent compared with the general situation every year. It can be concluded that part of countries from Asia and Africa bring down the benchmark. 
*I was in trouble to show the variable name of x-axis completely.*

The linear plot below is to show how the number of countries change, whose life expectancy is less than the worldwide median for each continent.

```{r The number of low life expectancy countries}
df %>% 
  ggplot(aes(x = year, y = num, color = continent)) +
  geom_line() +
  geom_point()
  
```

Except Oceania, the number of countries with low life expectancy has seen a decrease in every continent over time.

- **I want to do more**

Go on with the last task. To make it more meaningful, I decided to use yearly benchmark for measure, therefore it came out a different conclusion.

```{r}
# compute the median for each year
ww <- gapminder %>% 
  group_by(year) %>% 
  arrange(year) %>% 
  summarise(md = median(lifeExp)) %>% 
  `[[`("md")
# repeat the median for all countries
df <- gapminder %>% 
  mutate(md = rep(ww, nlevels(country))) 
# count the number of countries on each continent have a life expectancy less than the median, for each year
df %>% 
  filter(lifeExp < md) %>% 
  group_by(year, continent) %>%
  count(continent) %>% 
  knitr::kable()
```

```{r 2: The number of countries with low life expectancy }
df %>% 
  filter(lifeExp < md) %>% 
  group_by(year, continent) %>%
  count(continent) %>% 
  ggplot(aes(x = year, y = n, color = continent)) +
  geom_line() +
  geom_point()
  
```

Different from the last task, the number of countries with low life expectancy is basically flat  over time among four continents. There is a little decrease in Asia.
